---
title: "One with many reciprocal"
author: "Avi Kluger"
date: "January 18, 2019"
output: html_document
number_sections: true
---
##Summary of MLM results regarding errors, residuals, and covariances
package | $\sigma$^2^Focal 	|	$\sigma$^2^Partner	|$\sigma$^2^Residual	|Total $\sigma$^2^Focal	|Total $\sigma$^2^Partner|Cov (SE)| *r*(SE)
---- | ----- |---- | ----- |---- | ----- |---- | ----- |
SPSS|.549|.423|NA|.549|.423|NA|.239(.046)
lme4|.326|.200|.223|.549|.423|0.115(^1^)|.451
nlme|.376|.502|.046|.549|.423|^1^|.265
 |  | |  | |  | |  |
^1^cannot get confidence intervals on var-cov components: Non-positive definite approximate variance-covariance



```{r echo=FALSE, message=FALSE}
# Read (in SPSS format) from Kenny's book site and replicate Table 9.1
if (!require('foreign')) install.packages('foreign'); library('foreign')
table9.1_df <- read.spss("http://davidakenny.net/kkc/c9/4person.sav", 
               to.data.frame = TRUE, use.value.labels = FALSE)

if (!require("lavaan")) install.packages("lavaan"); 
suppressPackageStartupMessages(library(lavaan))
```


## Replicating SPSS MLM output with SEM
SPSS is taken from book's site. Results are identical to SPSS example in book's site and similar to p. 292


```{r}
# Define an indsitguisable model

OneWithManyIndistinguisableImitateSPSS <- '
# Actor effect:
      Actor   =~ 1*mfanx + 1*mcanx + 1*myanx

# Partner effect:
      Partner =~ 1*fmanx + 1*cmanx + 1*ymanx

# Relationship effects:
      # relationship.cm =~ 1* cmanx
      # relationship.fm =~ 1* fmanx
      # relationship.mc =~ 1* mcanx
      # relationship.mf =~ 1* mfanx
      # relationship.my =~ 1* myanx
      # relationship.ym =~ 1* ymanx

# Relationship :
       myanx ~~ va* myanx
       mfanx ~~ va* mfanx
       mcanx ~~ va* mcanx
       cmanx ~~ vp* cmanx
       fmanx ~~ vp* fmanx
       ymanx ~~ vp* ymanx

# Generalized reciprocity:
      Actor   ~~ Partner

# Dyadic reciprocity:
      mfanx  ~~  dr*fmanx 
      mcanx  ~~  dr*cmanx
      myanx  ~~  dr*ymanx

# Variance labels

      Actor   ~~ Actor 
      Partner ~~ Partner 

      # relationship.mc ~~ rva *  relationship.mc
      # relationship.mf ~~ rva *  relationship.mf
      # relationship.my ~~ rva *  relationship.my
      # relationship.cm ~~ rvp *  relationship.cm
      # relationship.fm ~~ rvp *  relationship.fm
      # relationship.ym ~~ rvp *  relationship.ym

# Intercepts

    mcanx ~ ia * 1
    mfanx ~ ia * 1
    myanx ~ ia * 1

    cmanx ~ ip * 1
    fmanx ~ ip * 1
    ymanx ~ ip * 1
'


# Estimate the model 
fitOneWithManyIndistinguisableImitateSPSS <- 
                 sem(OneWithManyIndistinguisableImitateSPSS, 
                 data       = table9.1_df,  
                 orthogonal = TRUE,
                 mimic      = "EQS")
# Examine the model.
summary(fitOneWithManyIndistinguisableImitateSPSS, fit.measures = TRUE, 
        standardized = TRUE)
```

###Use fSRM syntax.  
Took the fSRM syntax and deleted all elements that do not pertain to mother as to create one-with-many-reciprocal design.  Note that *lavaan* issues a warning "variables (theta) is not positive definite".  Dyadic reciporcity correlation is not calculated (Inf).  Problems seems similar to problem encountered with *lme4*.  However, estimates are different from *lme4*.  Note that the variances of the observed variables is forced to be zero, and the variance is "transfered" to the error.

```{r}
OneWithManyIndistinguisableImitate_fSRM <- '
# Actor effects:
    mActor   =~ 1*mfanx + 1*mcanx + 1*myanx
    
# Partner effects:
    mPartner =~ 1*fmanx + 1*cmanx + 1*ymanx
    
# Relationship effects:
    relationship.mc =~ 1* mcanx
    relationship.mf =~ 1* mfanx
    relationship.my =~ 1* myanx

    relationship.cm =~ 1* cmanx
    relationship.fm =~ 1* fmanx
    relationship.ym =~ 1* ymanx
    
# Fix observed variance to zero so it is "transfered" to the latent variables
    mcanx ~~ 0* mcanx
    mfanx ~~ 0* mfanx
    myanx ~~ 0* myanx

    cmanx ~~ 0* cmanx
    fmanx ~~ 0* fmanx
    ymanx ~~ 0* ymanx
    
# Generalized reciprocity:
    mActor   ~~ gr.m*mPartner
    
# Dyadic reciprocity:
    mfanx  ~~  dr*fmanx 
    mcanx  ~~  dr*cmanx
    myanx  ~~  dr*ymanx
    
# Variance labels
    mActor ~~ vam*mActor 
    mPartner ~~ vpm*mPartner 

# Constrain relationship variance within role to equality
    relationship.mc ~~ rv.a *  relationship.mc
    relationship.mf ~~ rv.a *  relationship.mf
    relationship.my ~~ rv.a *  relationship.my

    relationship.cm ~~ rv.p *  relationship.cm
    relationship.fm ~~ rv.p *  relationship.fm
    relationship.ym ~~ rv.p *  relationship.ym
'
# Estimate the model 
fitOneWithManyIndistinguisableImitate_fSRM <- 
                 sem(OneWithManyIndistinguisableImitate_fSRM, 
                 data = table9.1_df,  
                 orthogonal = TRUE,
                 mimic = "EQS")

# Examine the model.
summary(fitOneWithManyIndistinguisableImitate_fSRM, 
        fit.measures = TRUE, standardized=TRUE)
```

###Take the SPSS-compatible model and add intercepts. 
Constrain intercepts to equality with roles.  Note that the variance is now split between the observed variables (error in MLM?) and the residual (residual in MLM?).  The sums of va + vr and vp + vr are identical to the SPSS output.  The distribution of the variances differ from both *lme4* and *nlme*.
```{r}

OneWithManyIndistinguisableModelIntercepts <- '
# Actor effect:
      Actor   =~ 1*mfanx + 1*mcanx + 1*myanx

# Partner effect:
      Partner =~ 1*fmanx + 1*cmanx + 1*ymanx

# Relationship effects:
      relationship.cm =~ 1* cmanx
      relationship.fm =~ 1* fmanx
      relationship.mc =~ 1* mcanx
      relationship.mf =~ 1* mfanx
      relationship.my =~ 1* myanx
      relationship.ym =~ 1* ymanx

# Relationship :
       myanx ~~ va* myanx
       mfanx ~~ va* mfanx
       mcanx ~~ va* mcanx
       cmanx ~~ vp* cmanx
       fmanx ~~ vp* fmanx
       ymanx ~~ vp* ymanx

# Generalized reciprocity:
      Actor   ~~ Partner

# Dyadic reciprocity:
      mfanx  ~~  dr*fmanx 
      mcanx  ~~  dr*cmanx
      myanx  ~~  dr*ymanx

# Variance labels

      Actor   ~~ Actor 
      Partner ~~ Partner 

      relationship.mc ~~ rva *  relationship.mc
      relationship.mf ~~ rva *  relationship.mf
      relationship.my ~~ rva *  relationship.my
      relationship.cm ~~ rvp *  relationship.cm
      relationship.fm ~~ rvp *  relationship.fm
      relationship.ym ~~ rvp *  relationship.ym

# Intercepts

    cmanx ~ ip * 1
    fmanx ~ ip * 1
    mcanx ~ ia * 1
    mfanx ~ ia * 1
    myanx ~ ia * 1
    ymanx ~ ip * 1
'


# Estimate the model 
fitOneWithManyIndistinguisableModelIntercepts <- 
                 sem(OneWithManyIndistinguisableModelIntercepts, 
                 data = table9.1_df,  
                 orthogonal = TRUE,
                 mimic = "EQS")

# Examine the model.
summary(fitOneWithManyIndistinguisableModelIntercepts, 
        fit.measures = TRUE, standardized=TRUE)
```

###Constrain the residuals of the focal person and partners to equality.
This model is the same as the above, but produces one estimate of residual.  This constraint lead to a gain of one df.  The  The number of parameters is identical to MLM in R.  The various parameters related to the error are not identical $\chi$^2^.
```{r}
OneWithManyIndistinguisableEqualResiduals <- '
# Actor effect:
      Actor   =~ 1*mfanx + 1*mcanx + 1*myanx

# Partner effect:
      Partner =~ 1*fmanx + 1*cmanx + 1*ymanx

# Relationship effects:
      relationship.cm =~ 1* cmanx
      relationship.fm =~ 1* fmanx
      relationship.mc =~ 1* mcanx
      relationship.mf =~ 1* mfanx
      relationship.my =~ 1* myanx
      relationship.ym =~ 1* ymanx

# Relationship :
       myanx ~~ va* myanx
       mfanx ~~ va* mfanx
       mcanx ~~ va* mcanx
       cmanx ~~ vp* cmanx
       fmanx ~~ vp* fmanx
       ymanx ~~ vp* ymanx

# Generalized reciprocity:
      Actor   ~~ Partner

# Dyadic reciprocity:
      mfanx  ~~  dr*fmanx 
      mcanx  ~~  dr*cmanx
      myanx  ~~  dr*ymanx

# Variance labels

      Actor   ~~ Actor 
      Partner ~~ Partner 

      relationship.mc ~~ rv *  relationship.mc
      relationship.mf ~~ rv *  relationship.mf
      relationship.my ~~ rv *  relationship.my
      relationship.cm ~~ rv *  relationship.cm
      relationship.fm ~~ rv *  relationship.fm
      relationship.ym ~~ rv *  relationship.ym

# Intercepts

    cmanx ~ ip * 1
    fmanx ~ ip * 1
    mcanx ~ ia * 1
    mfanx ~ ia * 1
    myanx ~ ia * 1
    ymanx ~ ip * 1
'  
# Estimate the model 
fitOneWithManyIndistinguisableEqualResiduals <- 
                 sem(OneWithManyIndistinguisableEqualResiduals, 
                 data = table9.1_df,  
                 orthogonal = TRUE,
                 mimic = "EQS")

# Examine the model.
summary(fitOneWithManyIndistinguisableEqualResiduals, 
        fit.measures = TRUE, standardized=TRUE)
anova(fitOneWithManyIndistinguisableModelIntercepts,
  fitOneWithManyIndistinguisableEqualResiduals)